---
title: "Stat 302, Chapter 1 R demo"
author: "Brad McNeney"
output: html_document 
editor_options: 
  markdown: 
    wrap: 72
---

## RStudio and RMarkdown–

-   It is recommended that you work with R *via* RMarkdown, which allows
    you to intersperse text and R commands. When you "knit" the
    document, you see the text and output of your commands.
-   Topics for getting started with RMarkdown.
    -   Basics of RMarkdown
    -   Code chunks
    -   Sending code chunks to the R console
-   I will try to give you a template file to start for all the
    exercises.

## Jupyter notebooks

-   Some students prefer to use Jupyter *via* the online platform
    <https://sfu.syzygy.ca/> for analyses. For the most part, .Rmd files
    will look and run OK in Jupyter, but you may have to install R
    packages such as `Stat2Data` yourself. You should be able to do this
    by adding a code chunk with the following line to your notebook
    somewhere *before* you try to load `Stat2Data`.

```         
install.packages("Stat2Data",repos="https://muug.ca/mirror/cran/")
```

## Data for this demo

-   I'll illustrate with the `PorschePrice`, `SpeciesArea` and
    `Caterpillars` datasets from the `Stat2Data` R package.

```{r}
library(Stat2Data)
data("PorschePrice")
head(PorschePrice)
help(PorschePrice)
data("SpeciesArea")
head(SpeciesArea)
help(SpeciesArea)
data("Caterpillars")
```

## R functions

-   A function has input and output.
-   For example, the `head()` function takes a dataset as input and
    returns a printout of the top few (by default 6) lines of the
    dataset.
-   Function outputs can be text printed to the R console, plots sent to
    the Plots tab, help pages sent to the Help tab, spreadsheets that
    open as new tabs, etc.
    -   E.G., To see the entire dataset in spreadsheet form use
        `View(PorschePrice)`.

## Plotting data

-   There is a `plot()` function in R that does high-quality graphical
    displays of data, but the `ggplot()` function from the `ggplot2`
    package does better.

```{r}
library(tidyverse) # includes the ggplot2 package for plotting data
ggplot(PorschePrice,aes(x=Mileage,y=Price)) + 
  geom_point() +
  geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
```

## Adding or modifying variables

-   Recall from lecture that the plot of `Species` *versus* `Area`
    (`SpeciesArea` dataset) was not very informative.
-   This dataset also includes the natural log of the two variables.
-   Statisticians like natural log, but most people prefer log-base-10.
    -   A one unit increase on the log-10 scale is a 10-fold increase on
        the original scale.
-   Add new variables to a dataset with the `mutate()` function (from
    tidyverse)

```{r}
SpeciesArea <- mutate(SpeciesArea, 
                      log10Area=log10(Area), log10Species=log10(Species))
ggplot(SpeciesArea,aes(x=log10Area,y=log10Species)) + 
  geom_point() + geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
```

-   The `Caterpillars` dataset has a variable `Instar` that codes the
    life stage of each caterpillar. This is a categorical variable, but
    R thinks it is quantitative.
-   We change to a categorical, or factor variable as follows.

```{r}
Caterpillars <- mutate(Caterpillars,Instar = factor(Instar))
```

## Least squares regression

-   We fit models with the `lm()` function (linear model).
-   Models are specified by formulas with the response variable on the
    left and the explanatory variables on the right of a tilde (`~`)
    sign.
-   We also need to input the dataset name.

```{r}
pfit <- lm(Price ~ Mileage, data=PorschePrice)
pfit
```

-   The assignment `<-` saves the fitted model for later use.
    -   The name `pfit` is arbitrary.
    -   We'll call the saved output an "object".

## Fitted model summary

-   To get a basic summary of the model, just type the object name into
    the R console.

```{r}
pfit
```

## Checking model assumptions - residuals versus fitted values

```{r, fig.height=3, fig.width=5}
plot(pfit,which=1)
```

## Checking model assumptions - Normal quantile plot

```{r}
plot(pfit,which=2)
```

## Non-linearity in a residual plot

```{r}
simdat <- function(n) {
  x <- 3+ rnorm(n) 
  y <- 1 + 2*x + x^2 + rnorm(n) 
  data.frame(x=x,y=y)
}
plotres <- function(dat){
  ff <- lm(y~x,data=dat) 
  plot(ff,which=1)
}

dat <- simdat(100)
ggplot(dat,aes(x=x,y=y)) + geom_point() + geom_smooth() +
  geom_smooth(method="lm",se=FALSE,color="red")
plotres(dat)

dat <- simdat(100)
plotres(dat)

dat <- simdat(100)
plotres(dat)

dat <- simdat(100)
plotres(dat)
```

## Non-constant SD in a residual plot

```{r}
simdat <- function(n) {
  x <- 3+ rnorm(n) 
  y <- 1 + 2*x  + x* rnorm(n) 
  data.frame(x=x,y=y)
}

dat <- simdat(100)
ggplot(dat,aes(x=x,y=y)) + geom_point() + geom_smooth() +
  geom_smooth(method="lm",se=FALSE,color="red")
plotres(dat)

dat <- simdat(100)
plotres(dat)

dat <- simdat(100)
plotres(dat)

dat <- simdat(100)
plotres(dat)
```

## Non-normal errors in a Q-Q plot

```{r}
simdat <- function(n) {
  x <- 3+ rnorm(n) 
  y <- 1 + 2*x  + rexp(n) 
  data.frame(x=x,y=y)
}

dat <- simdat(100)
plotQQ <- function(dat){
  ff <- lm(y~x,data=dat) 
  plot(ff,which=2)
}
plotQQ(dat)

dat <- simdat(100)
plotQQ(dat)

simdat <- function(n) {
  x <- 3+ rnorm(n) 
  y <- 1 + 2*x  - rexp(n) 
  data.frame(x=x,y=y)
}

dat <- simdat(100)
plotQQ(dat)

dat <- simdat(100)
plotQQ(dat)
```
