---
title: "Chapter 3 R demo 1"
author: "Brad McNeney"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data

```{r, warning=FALSE, message=FALSE}
library(Stat2Data)
library(tidyverse)
data(Caterpillars)
data(SpeciesArea)
```

-   Remember that you can type `help("Caterpillars")` or
    `help("SpeciesArea")` for information about these datasets and
    `View(Caterpillars)` or `View(SpeciesArea)` to view them.

## Multiple regression for the Caterpillars data

-   We will do a regression of `LogMass` on `LogWetFrass` and `Fgp`.
-   consider we could do one line for red data one line for blue data
    -   BUT THIS IS NOT THE MODEL DISCUSSED IN LECTURE
    -   add this to code to see:
        -   geom_smooth(method = "lm")
-   for the model we fit we would expect the red line and red line to
    run parallel, always at the same distance apart
    -   in this example the distance between the lines should be:

        ```         
        0.596424
        ```

```{r}
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Fgp)) + geom_point() # + geom_smooth(method = "lm")
```

-   Fit a regression model for the Caterpillars data that includes an
    effect for `Fgp` as follows.

```{r}
cfit <- lm(LogWetFrass ~ LogMass + Fgp,data=Caterpillars)
```

-   Extract the coefficients table from the model summary:

```{r}
summary(cfit)$coefficients
```

-   The full model summary includes the following model summaries:
    -   The `Residual standard error` is the estimate
        $\sqrt{SSE/(n-k-1)}$ of the error SD in the model.
    -   The `F statistic` and p-value are from the F test.
    -   `Multiple R-squared` is the $R^2$.

```{r}
summary(cfit)
```

-   Other "extractor" functions we discussed for linear regression work
    the same:
-   residual vs fitted looks slightly better when we include free growth
    period (Fgp)
-   something going on with Q-Q plot, shows normal dist from -1 upwards,
    residuals too low at -4, heavy left tail for that residual
    distribution
-   Cook's distance is gud (nothing greater than 1)

```{r}
confint(cfit,level=0.95)
plot(cfit,which=1) 
plot(cfit,which=2)
plot(cfit,which=5)
Caterpillars <- mutate(Caterpillars,residuals=residuals(cfit))
```

-   Another useful extractor function is `predict`.
    -   We used this in Chapter 2 to get predictions at new points plus
        confidence/prediction intervals.
    -   By default, `predict()` returns the fitted values, $\hat{y}$,
        for all the observed data points. This can be useful for
        plotting the fitted model.
    -   We use `geom_line()` but modify the context: for the line map
        the $\hat{y}$s to the y-axis of the plot (but use the same x's
        as for the points).
    -   Verify that the vertical distance between the two lines is equal
        to the `Fgp` coefficient $0.59642$; e.g., at `LogMass`=0.

#predict(cfit) is all of the predicted y hat values for each of the
caterpillars

#geom_line(aes(y=yhats) create lines using the y hat values (i.e. the
predicted LogWetFrass values)

```{r}
Caterpillars <- mutate(Caterpillars,yhats=predict(cfit))
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Fgp)) + 
  geom_point() + geom_line(aes(y=yhats))
```

## Multiple regression for the SpeciesArea dataset

```{r}
ggplot(SpeciesArea,aes(x=logArea,y=Species)) + geom_point() + geom_smooth()
```

-   A cubic polynomial might fit. Let's compare this to the scatterplot
    smoother.

CUBIC THINK BUMP DIP OR DIP BUMP

#ggplot(SpeciesArea,aes(x=logArea,y=Species)) + geom_point() +
geom_smooth()

–\> fitter smoother fn

#ggplot(SpeciesArea,aes(x=logArea,y=Species)) + geom_point() +
geom_smooth(method="lm",formula=y\~poly(x,3),color="red")

--\> fit with 3rd degree polynomial

y = b0 + b1\*x + b2\*x\^2

```{r}
ggplot(SpeciesArea,aes(x=logArea,y=Species)) + geom_point() + 
  geom_smooth() + geom_smooth(method="lm",formula=y~poly(x,3),color="red")
```

-   Fit the cubic polynomial to the data.
    -   More on polynomial regression in Section 3.4 of the text.
-   multiple regression can be used to fit cubics

```{r}
sfit <- lm(Species ~ poly(logArea,3),data=SpeciesArea)
summary(sfit)
```
