---
title: "Chapter 3, R demo 2"
author: "Brad McNeney"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Confounding demo

-   Simulate some data to illustrate confounding.
-   The explanatory variable of interest is $X_1$ and the extraneous
    variable is $X_2$.

```{r}
set.seed(1)
X2 <- c(rep(0,50),rep(1,50))
X1 <- c(rnorm(50,mean=-1),rnorm(50,mean=1))
Y <- rnorm(100,mean=5+X1-5*X2)
dat <- data.frame(Y,X1,X2)
```

-   To be a confounder, $X_2$ must be associated with both $Y$ and
    $X_1$.

```{r}
cor(dat)
```

-   Fit models with and without $X_2$.

```{r}
summary(lm(Y~X1,data=dat))$coefficients
summary(lm(Y~X1+X2,data=dat))$coefficients
abs(-0.326 - 0.998)/abs(0.998) * 100 # 133 percent change
```

-   Now visualize the data to see why the absence of the confounder
    gives a misleading estimate of the relationship between $X_2$ and
    $Y$.

-   this plot tells us that X2 has a strong effect on Y

    -   if you fix X1 and take the distance between the two lines

-   strong correlation between X1 and X2

-   can think of this plot as pulling apart correlation between X1 and Y
    (up and down), and between X2 and Y (left and right)

-   if you did least squares regression on the data without considering
    X2 you would expect the resulting line to have a negative slope –\>
    dramatic change!!!

    -   ignoring X2 gives us the wrong conclusion of X1 effect on Y

-   hierarchy check if:

    -   modifier
    -   confounder

```{r}
library(tidyverse)
dat <- mutate(dat,X2 = factor(X2))
ggplot(dat,aes(y=Y,x=X1,color=X2)) + 
  geom_point() +
  geom_smooth(method="lm")
```

## The `Caterpillars` data

```{r}
library(Stat2Data)
data(Caterpillars)
```

-   Type `help(Caterpillars)` for information about the dataset and
    `View(Caterpillars)` to view it.

## Multiple linear regression with the `Caterpillars` data

-   Fit the regression model with interaction between `LogMass` and
    `Fgp`.
-   The model is specified by the "formula" `LogWetFrass ~ LogMass*Fgp`,
    which translates to fitting

$$
{\rm LogWetFrass} = \beta_0 + \beta_1 {\rm LogMass} + 
\beta_2 {\rm FgpY} +  \beta_3 {\rm LogMass}\times {\rm FgpY} + \epsilon.
$$

where `FgpY` is a "dummy variable" coded as 1 for yes and 0 for no.

-   \* different intercepts and slopes

-   \+ different intercepts

-   start with \* and then we will deem which is appropriate

-   mutate(Caterpillars,yhats_int=predict(cfit_int))

    -   adding y hats to model so we can plot the two separate lines

-   geom_line(aes(y=yhats_int))

    -   draw the lines

```{r}
cfit_int <- lm(LogWetFrass ~ LogMass * Fgp,data=Caterpillars)
summary(cfit_int)$coefficients
Caterpillars <- mutate(Caterpillars,yhats_int=predict(cfit_int))
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Fgp)) +
  geom_point() + geom_line(aes(y=yhats_int))
```

-   The model diagnostic plots are obtained as before.
    -   Still not great - QQ plot seems to have a trend
    -   no influential observations

```{r}
plot(cfit_int,which=1)
plot(cfit_int,which=2)
plot(cfit_int,which=5)
```

-   To improve the fit we could try using the life stage variable
    `Instar`.
-   `Instar` is a categorical variable coded from 1 to 5. Will be
    interpreted as numeric in R, but we want it to be a "factor" for
    modelling.
    -   Instar = factor(Instar)
    -   replace numeric variable with a factor/categorical variable

```{r}
Caterpillars <- mutate(Caterpillars,Instar = factor(Instar))
```

```{r}
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Instar)) +
  geom_point() + 
  geom_smooth() + 
  geom_smooth(method="lm")
```

-   Fit the regression model with "main effects" for `LogMass` and
    `Instar` and obtain a summary of the regression coefficients as
    follows.

+only allows fo different intercepts

```{r}
ff <- lm(LogWetFrass ~ LogMass + Instar,data=Caterpillars)
summary(ff)$coefficients
confint(ff,level=0.95)
```

-   The model is specified by the formula
    `LogWetFrass ~ LogMass + Instar`, which translates to fitting

$$
{\rm LogWetFrass} = \beta_0 + \beta_1 {\rm LogMass} + 
\beta_2 {\rm Instar2} + \ldots + \beta_5 {\rm Instar5} + \epsilon.
$$

-   The variable `Instar2` takes value 1 if category 2 and 0 otherwise.

-   The variables `Instar3`, ..., `Instar5` are analogous.

-   R creates these "dummy" variables for us because `Instar` is a
    factor.

-   To add interaction between `LogMass` and `Instar`, use `*`.

```{r}
ff <- lm(LogWetFrass ~ LogMass * Instar,data=Caterpillars)
summary(ff)$coefficients
```

-   **Exercise:** Do the diagnostic plots for the model fit above. Does
    the QQ-plot look any better with this model than the one with `Fgp`?

-   For predictions we need to specify values of **both** explanatory
    variables.

    -   Make sure the variable names in your new dataset match those in
        the formula **exactly**.

-   create new dataset for prediction intervals

    -   nn \<- data.frame(LogMass=0,Instar="4")

    -   bc instar is categorical must "4"

```{r}
nn <- data.frame(LogMass=0,Instar="4")
predict(ff,newdata=nn,interval="confidence")
```

-   What about different lines for each *combination* of `Instar` and
    `Fgp`?

```{r}
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Instar,shape=Fgp)) +
  geom_point() + 
  geom_smooth(method="lm")
```

-   statistical interaction between all three variables:

    -   lm(LogWetFrass \~ LogMass \* Instar \* Fgp,data=Caterpillars)

    -   baseline group: fgp = No, instar = "1"

    -   

```{r}
ffFgp <- lm(LogWetFrass ~ LogMass * Instar * Fgp,data=Caterpillars)
summary(ffFgp)$coefficients
```

```{r}
plot(ffFgp,which=1)
plot(ffFgp,which=2)      #still have a heavy left tail :()
```

-   We can get quite creative with the multiple regression models we
    fit, but how do we know how complicated to make the model?
    -   We will study methods to test subsets of predictors in Section
        3.6, and more on model selection in Chapter 4.

```{r}
anova(ff,ffFgp)
```
