---
title: "Stat 302, Test your R installation"
author: "Brad McNeney"
output: html_document 
editor_options: 
  markdown: 
    wrap: 72
---

## RStudio and RMarkdown

-   It is recommended that you work with R *via* RMarkdown, which allows
    you to intersperse text and R commands. When you "knit" the
    document, you see the text and output of your commands.
-   I will give you a template file to start for all the exercises.

## Jupyter notebooks

-   Some students prefer to use Jupyter *via* the online platform
    <https://sfu.syzygy.ca/> for analyses. For the most part, .Rmd files
    will look and run OK in Jupyter, but you may have to install R
    packages such as `Stat2Data` yourself. You should be able to do this
    by adding a code chunk with the following line to your notebook
    somewhere *before* you try to load `Stat2Data`.

```         
install.packages("Stat2Data",repos="https://muug.ca/mirror/cran/")
```

## Data for this test

-   We'll test with the `PorschePrice` dataset from the `Stat2Data` R
    package.

```{r}
library(Stat2Data)
data("PorschePrice")
head(PorschePrice)
```

## Test code

-   If you have installed everything correctly, you should be able to
    run the following code chunk.
-   To run a code chunk, you can
    -   click the "play button" at the top-right of the chunk, or
    -   put your cursor on the first line of code and type ctrl-Return
        (hold the control key down and then hit the Return key) on each
        line to run the lines one-by-one.

```{r}
library(tidyverse)
ggplot(PorschePrice,aes(x=Mileage,y=Price)) + 
  geom_point() +
  geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
```

-   If you see an error that indicates `tidyverse` or `ggplot` are
    missing, find the `Tools -> Install Packages` menu, enter
    `tidyverse` in the Packages text field of the pop-up window, and
    click `Install`.

## Knitting the document

-   To run all code chunks and produce an html output file with all of
    the R output, click the "knit" button in the toolbar just above the
    contents of the file. It has an icon that looks like a ball of wool
    with a knitting needle in it, and may also have the word "Knit"
    beside it.

-   Try knitting the document to make sure your installation is
    complete.
