---
title: "Chapter 1 practice exercises"
author: "Brad McNeney"
date: '2025-01-10'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

-   These exercises are adapted from the Stat2 course text.

-   If you are using Jupyter, add a chunk to your notebook with the
    following line *before* you try to load `Stat2Data`.

```         
install.packages("Stat2Data",repos="https://mirror.rcg.sfu.ca/mirror/CRAN/")
```

-   Now load all the R packages we'll need. Remember to do this for
    every R session.

```{r, warning=FALSE}
library(Stat2Data)
library(tidyverse)
```

## 1

Consider a fitted regression equation $\hat{Y} = 100 + 15 X$. Which of
the following is false?

a.  The fitted slope is 15.
b.  The predicted value of $Y$ when $X=0$ is 100.
c.  The predicted value of $Y$ when $X=2$ is 110. **–\> FALSE**
d.  Larger values of $X$ are associated with larger values of $Y$.

## 2

Data on 116 Savannah sparrows includes measurments of weight and wing
length. The fitted least-squares regression model is summarized below.

```{r}
data("Sparrows")
lm(Weight ~ WingLength,data=Sparrows)
```

a.  What is the slope of the least squares regression line?

    **0.4674**

b.  What is the intercept of the least squares regression line?

    **1.3655**

## 3

Consider the fitted regression equation $Y = 25 + 2 X$. If $x_1=13$ and
$y_1 = 50$, what is the residual for the first data point?

**-1**

## 4

Which of the following regression assumptions does the plot of residuals
*versus* fitted values **not** assess?

a.  Linearity
b.  Constant variance
c.  **Independence**

## 5

The number of calories and number of grams of sugar per serving were
measured for 36 breakfast cereals. The following graph and fitted least
squares regression explore the relationship between these variables.

```{r}
data("Cereal")
ggplot(Cereal,aes(x=Sugar,y=Calories)) +
  geom_point() +  geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
ff <- lm(Calories ~ Sugar,data=Cereal)
ff
```

a.  Which variable is considered the response in the above plot and
    model?

    Calories

b.  Interpret the slope of the fitted model.

    **The slope is 2.481. It is positive suggesting the possibility of a
    positive association. If so, for every 1 g increase of sugar in
    cereal, the calories in the cereal increases by 2.48.**

c.  Mueslix has 13 grams of sugar. How many calories does the fitted
    model predict for this cereal?

    **119.7 calories**

d.  What is the residual for Mueslix?

    **Actual calories = 160, predicted calories = 119.7**

    **Residual = 160 - 119.7 = 40.3**

e.  Based on the plot produced by the above code chunk, does the linear
    regression model appear to be a good summary of the relationship
    between calories and sugar?

    **No by looking at the smooth scatter function (blue line) you can
    see there are two bumps in the data, suggesting a cubic model may
    summarize the relationship better.**

## 6

Type `help("SpeciesArea")` into the R console to read about the dataset
on islands in Southeast Asia. The following R code examines the
relationship between the area and number of mammal species on the
islands, both on the log-10 scale.

```{r}
data("SpeciesArea")
SpeciesArea <- mutate(SpeciesArea,
                      log10Area = log10(Area),log10Species=log10(Species))
ggplot(SpeciesArea,aes(x=log10Area,y=log10Species)) + 
  geom_point() +  geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
ff <- lm(log10Species ~ log10Area, data=SpeciesArea)
ff
plot(ff,which=1)
```

a.  Interpret the slope of the fitted model.

    **The slope is 0.2355. It is positive suggesting the possibility of
    a positive association between the X and Y variables. If so, for
    every a 1% increase in area, the number of mammal species increases
    by about 0.236%.**

My bad this is for if the y data is the only log transformed data

must exponentiate the slope to make it understandable.
exp(coef(ff)["log10Area]") = 1.265509

subtract by 1 and multiply by 100 to get one unit-increase info

(1.265509 - 1) \* 100 = 26.6%

This says every one-unit increase in x is multiplied by about 1.266. Or
in other words, for every one-unit increase in x, y increases by about
26%.

a.  Make a plot of the residuals versus fitted values and comment on the
    pattern.

    plot(ff,which=1)

    **Based on the residuals vs fitted values the data set does not
    appear to have a linear relationship but there are not many data
    points. The dip of the line indicates that a higher order polynomial
    may better fit the data set.**

## 7

Load the `Caterpillars` dataset into R and read its help file.

a.  Produce a scatterplot for predicting `WetFrass` with `Mass`.

b.  Produce a scatterplot for predicting the log-base-10 of `WetFrass`
    with the log-base-10 of `Mass`.

c.  Which of (a) and (b) are better summarized by a least squares
    regression?

    **(b) the log transformed data**

d.  The following R code produces a scatterplot with different colors
    for different life stages. Does the relationship between
    log-10-`WetFrass` and log-10-`Mass` look consistent across the
    different life stages?

    **The relationship does not appear to be consistent across the
    different life stages. The relationship appears to start off cubic
    then quadratic then more linear as you move left to right across the
    plot.**

```{r}
data("Caterpillars")
Caterpillars <- mutate(Caterpillars,Instar = factor(Instar))



# a)
ggplot(Caterpillars,aes(x=Mass,y=WetFrass)) +
  geom_point() + geom_smooth() +
  geom_smooth(method="lm", se=FALSE, color="red")
# b) 
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass)) +
  geom_point() + geom_smooth() +
  geom_smooth(method="lm", se=FALSE, color="red")



ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Instar)) +
  geom_point() + geom_smooth()
```
