---
title: "Chapter 1 practice exercises"
author: "Brad McNeney"
date: '2025-01-10'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

* These exercises are adapted from the Stat2 course text.

* If you are using Jupyter, add a chunk to your notebook 
with the following line
*before* you try to load `Stat2Data`.

```
install.packages("Stat2Data",repos="https://mirror.rcg.sfu.ca/mirror/CRAN/")
```

* Now load all the R packages we'll need. Remember
to do this for every R session.
```{r, warning=FALSE}
library(Stat2Data)
library(tidyverse)
```


## 1
Consider a fitted regression equation $\hat{Y} = 100 + 15 X$. Which
of the following is false?

a. The fitted slope is 15.
b. The predicted value of $Y$ when $X=0$ is 100.
c. The predicted value of $Y$ when $X=2$ is 110.
d. Larger values of $X$ are associated with larger values of $Y$.

## 2
Data on 116 Savannah sparrows includes measurments of weight and
wing length. The fitted least-squares regression model is summarized
below.

```{r}
data("Sparrows")
lm(Weight ~ WingLength,data=Sparrows)
```

a. What is the slope of the least squares regression line?
b. What is the intercept of the least squares regression line?

## 3
Consider the fitted regression equation $Y = 25 + 2 X$. If $x_1=13$
and $y_1 = 50$, what is the residual for the first data point?

## 4

Which of the following regression assumptions does the 
plot of residuals *versus* fitted values **not** assess?

a. Linearity
b. Constant variance
c. Independence


## 5
The number of calories and number of grams of sugar per serving were 
measured for 36 breakfast cereals. The following graph and 
fitted least squares regression explore the relationship 
between these variables.

```{r}
data("Cereal")
ggplot(Cereal,aes(x=Sugar,y=Calories)) +
  geom_point() +  geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
ff <- lm(Calories ~ Sugar,data=Cereal)
ff
```

a. Which variable is considered the response in the above plot and model?
b. Interpret the slope of the fitted model.
c. Mueslix has 13 grams of sugar. How many calories does 
the fitted model predict for this cereal?
d. What is the residual for Mueslix?
e. Based on the plot produced by the above code chunk, does
the linear regression model appear to be a good summary of the 
relationship between calories and sugar?

## 6

Type `help("SpeciesArea")` into the R console to read about the 
dataset on islands in Southeast Asia.
The following R code examines the relationship 
between the area and number of 
mammal species on the islands, both on the log-10 scale.

```{r}
data("SpeciesArea")
SpeciesArea <- mutate(SpeciesArea,
                      log10Area = log10(Area),log10Species=log10(Species))
ggplot(SpeciesArea,aes(x=log10Area,y=log10Species)) + 
  geom_point() +  geom_smooth() + 
  geom_smooth(method="lm",se=FALSE, color="red")
ff <- lm(log10Species ~ log10Area, data=SpeciesArea)
ff
```

a. Interpret the slope of the fitted model.
b. Make a plot of the residuals versus fitted values and comment 
on the pattern.

## 7
Load the `Caterpillars` dataset into R and read its help file.

a. Produce a scatterplot for predicting `WetFrass` with `Mass`. 
b. Produce a scatterplot for predicting the log-base-10 of
`WetFrass` with the log-base-10 of `Mass`. 
c. Which of (a) and (b) are better summarized
by a least squares regression?
d. The following R code produces a scatterplot with different 
colors for different life stages. Does the relationship
between log-10-`WetFrass` and log-10-`Mass` look 
consistent across the different life stages?

```{r}
data("Caterpillars")
Caterpillars <- mutate(Caterpillars,Instar = factor(Instar))
ggplot(Caterpillars,aes(x=LogMass,y=LogWetFrass,color=Instar)) +
  geom_point() + geom_smooth()
```
